%-------------------------------------------------------------------------------
% PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%-------------------------------------------------------------------------------

\documentclass[a4paper,10p,openright]{memoir}

\usepackage{dikuReport}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{latexsym}
\usepackage{caption}
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage{extramarks}
\usepackage{color}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{courier}
\usepackage{lipsum}
\usepackage{tabularx}
\usepackage{mathtools}
\usepackage{mathrsfs}
\usepackage{tikz}
\usepackage{lstrfun}
\usepackage{xspace}
\usepackage{cite}
\usepackage{microtype}
\usepackage{caption}
\usepackage{todonotes}
\usepackage{subcaption}
\usepackage{url}
\usepackage{stmaryrd}
\usepackage{ebproof}
\usepackage{xcolor,calc}
\usepackage{float}
\usepackage{morefloats}
\usepackage{latexsym}
\usepackage{fix-cm}
\usepackage{pdfpages}
\usepackage{amsbsy}
\usepackage{alltt}
\usepackage{relsize}
\usepackage{ifthen}
\usepackage[]{units}
\usepackage{hyperref}

% Include some layout setup.
\input{layout.tex}

\def\UrlFont{\rmfamily}

\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}

% Margins
% \setlength{\marginparwidth}{2.75cm}
\topmargin=0in
\evensidemargin=0.0in
\oddsidemargin=0.0in
\textwidth=6.2in
\textheight=8.5in
\headsep=0.35in

% \makeatletter
% \renewcommand\chapter{\if@openright\cleardoublepage\else\clearpage\fi
%                     \thispagestyle{fancy}
%                     \global\@topnum\z@
%                     \@afterindentfalse
%                     \secdef\@chapter\@schapter}
% \makeatother

\linespread{1.1} % Line spacing

% Set up the header and footer
% \pagestyle{fancy}
% \lhead{} % Top left header
% \chead{}
% \rhead{} % Top right header
% \lfoot{\lastxmark} % Bottom left footer
% \cfoot{} % Bottom center footer
% \rfoot{Page\ \thepage} % Bottom right footer
% \renewcommand\headrulewidth{0pt} % Size of the header rule
% \renewcommand\footrulewidth{0.4pt} % Size of the footer rule

% \setlength\parindent{0pt} % Removes all indentation from paragraphs

\tikzset{node distance=2cm, auto}

\lstset{language=RFun}

\newcommand*{\anddot}{%
  \mathclose{}%
  \nonscript\mskip.5\thinmuskip
  \boldsymbol{.}%
  \;%
  \mathopen{}%
}

\def\rfunc{\ensuremath{\mathsf{CoreFun}}\xspace}
% \def\rfunfst{\ensuremath{\mathsf{RFun}_{fst}}\xspace}
\def\rfun{\ensuremath{\mathsf{RFun}}\xspace}

%-------------------------------------------------------------------------------
% SEMANTICS COMMANDS
%-------------------------------------------------------------------------------

% Environments
\newcommand{\cod}{\text{cod}}
\newcommand{\dom}{\text{dom}}
\newcommand{\Hom}{\text{Hom}}

\newcommand{\leaves}{\text{leaves}}
\newcommand{\sfmp}{\text{staticFMP}}
\newcommand{\PLVal}{\text{PLVal}}

\def\vdashe{\vdash}
\def\vdashd{\vdash\!\!\!\vdash}
\newcommand{\con}{\Sigma;\Gamma\vdash}
\newcommand{\conp}{\Sigma;\Gamma'\vdash}
\newcommand{\conpp}{\Sigma;\Gamma''\vdash}
\newcommand{\conemp}{\Sigma;\emptyset\vdash}
\newcommand{\concup}{\Sigma;\Gamma\cup\Gamma'\vdash}
\newcommand{\concupp}{\Sigma;\Gamma\cup\Gamma''\vdash}
\newcommand{\pcon}{p \vdash}
\newcommand{\storeinv}{p;\sigma\vdash^{-1}}
\newcommand{\emptyinv}{p;\emptyset\vdash^{-1}}

\newcommand{\dcon}{\Delta\vdash}

% Constructs
\newcommand{\inl}[1]{\textbf{inl}(#1)}
\newcommand{\inr}[1]{\textbf{inr}(#1)}

\newcommand{\lett}[3]{\textbf{let } #1 = #2 \textbf{ in } #3}
\newcommand{\caseof}[5]{\textbf{case } #1 \textbf{ of } #2 \Rightarrow #3, #4 \Rightarrow #5}
\newcommand{\caseofu}[5]{\textbf{case } #1 \textbf{ of } #2 \Rightarrow #3, #4 \Rightarrow #5 \textbf{ unsafe }}
\newcommand{\caseofs}[6]{\textbf{case } #1 \textbf{ of } #2 \Rightarrow #3, #4 \Rightarrow #5 \textbf{ safe } #6}
\newcommand{\smpcase}[2]{\textbf{case } #1 \textbf{ of } #2}

\newcommand{\roll}[2]{\textbf{roll } [#1]\ #2}
\newcommand{\unroll}[2]{\textbf{unroll } [#1]\ #2}

\newcommand{\abort}[2]{\textbf{abort}_#1\ #2}

\newcommand{\class}[2]{\textbf{class } #1 \textbf{ where } #2}
\newcommand{\instance}[2]{\textbf{instance } #1 \textbf{ where } #2}

\newcommand{\within}[2]{\textbf{within } #1 : #2 \textbf{ end}}

\newcommand{\defeq}{\stackrel{\>\text{def}}{=}\>}

\begin{document}

%-------------------------------------------------------------------------------
% TITLE PAGE
%-------------------------------------------------------------------------------

% Basic information
\thesistype{MSc thesis}
\thesiscomment{} % You can leave this blank
\title{Design of a Reversible Functional Programming Language}
\subtitle{And its Type System}
\author{Petur Andrias Højgaard Jacobsen}
\cosupervisor{Robin Kaarsgaard Jensen}
\supervisor{Michael Kirkedal Thomsen}
\date{\today} % Hand-in date
%\subject{The short description that is suitable for a database.} % This is not needed.

% Make the front page, title page, and other required information.
\pagestyle{plain}
\maketitle

% Start at page 3. I do not count the front page in the numbering.
\cleardoublepage
\pagenumbering{roman}
\setcounter{page}{3}

\cleardoublepage
\pagestyle{plain}

%-------------------------------------------------------------------------------

\begin{abstract}

  Reversible programming languages are languages which exhibit both forward and
  backward determinism. The theory of reversible flowchart languages is
  relatively well understood, but studies of reversible functional languages
  are few and far between. In this thesis, we introduce a garbage-free,
  reversible, function language, inspired by \rfun, which we call \rfunc. We
  provide a formalization of its semantics and extend it with a type system
  based on relevance logic. The type system also has support for recursive and
  polymorphic types. With the type system, we are able to add the use of
  ancillae variables through an unrestricted fragment. Backwards determinism of
  of branching is achieved with a First Match Policy, but we investigate the
  possibility of ensuring backwards determinism with exit assertions or static
  guarantees of orthogonality as alternatives. As program inversion of
  non-flowchart languages generally is hard, we present a formalization of its
  inverse semantics instead. Finally, we describe how to lift the core language
  into a syntactically lighter language via a sequence of translation schemes,
  making it more amenable for modern style programming.

\end{abstract}

% Table of contents
\cleardoublepage
\chapterstyle{combined}
\tableofcontents*

% Starting the real text.
\cleardoublepage
\pagenumbering{arabic}
\setcounter{page}{1}

%-------------------------------------------------------------------------------

\chapter{Introduction}

Reversible computing is the study of computational models in which individual
computation steps can be uniquely and unambiguously inverted. For programming
languages, this means languages in which programs can be run \emph{backward}
and get a unique result (the exact input). In this thesis, we restrict ourselves
to \emph{garbage-free} reversible programming languages, which guarantee not
only that all programs are reversible, but also that no hidden duplication of
data is required in order to make this guarantee.

In this thesis, we present a simple, but \emph{r-Turing
complete}~\cite{AxelsenGlueck:2016}, reversible typed functional programming
language, \rfunc. Functional languages and programming constructs are currently
quite successful; this includes both applications in special domains, e.g.
Erlang, and functional constructs introduced in mainstream programming
languages, such as Java and C++. We believe that functional languages also
provide a suitable environment for studying reversible programs and
computations, as recently shown in~\cite{ThomsenAxelsen:2016:IFL}. However, the
lack of a type system exposed the limitations of the original \rfun language,
which has motivated this work. A carefully designed type system can provide
better handling of static information through the introduction of
\emph{ancillae typed} variables, which are guaranteed to be unchanged across
function calls. Further, it can often be used to statically verify the
\emph{first match policy} that is essential to reversibility of partially
defined functions. It should be noted that this type system is not meant to
guarantee reversibility of well-typed programs (rather, guaranteeing
reversibility is a job for the \emph{semantics}). Instead, the type system aids
in the clarity of expression for programs, provides fundamental
well-behavedness guarantees, and is a source of additional static information
which can enable static checking of certain properties, such as the
aforementioned \emph{first-match policy}.

An implementation of the work in this thesis can be found at:

\begin{center}
  \texttt{\url{https://github.com/diku-dk/coreFun/}}
\end{center}

\section{Background}

Initial studies of reversible (or information lossless) computation date back
to the years around 1960. These studies were based on quite different
computation models and motivations: Huffman studied information lossless finite
state machines for their applications in data transmission~\cite{Huffman:1959},
Landauer came to study reversible logic in his quest to determine the sources
of energy dissipation in a computing system~\cite{Landauer:1961}, and Lecerf
studied reversible Turing machines for their theoretical
properties~\cite{Lecerf:1963}.

Although the field is often motivated by a desire for energy and entropy
preservation though the work of Landauer~\cite{Landauer:1961}, we are more
interested in the possibility to use reversibility as a property that can aid
in the execution of a system, an approach which can be credited to
Huffman~\cite{Huffman:1959}. It has since been used in areas like programming
languages for quantum computation~\cite{GreenEtAl:2013:Quip}, parallel
computing~\cite{SchordanEtal:2015:RC}, and even
robotics~\cite{SchultzLEA:2015}. This diversity motivates studying reversible
functional programming (and other paradigms) independently, such that we can
get a better understanding of how to improve reversible programming in these
diverse areas.

The earliest reversible programming language (to the authors' knowledge) is
Janus, an imperative language invented in the 1980's, and later
rediscovered~\cite{LutzDerby:1986,YokoyamaGlueck:2007:Janus} as interest in
reversible computation spread. Janus and languages deriving from it have
since been studied in detail, so that we today have a reasonably good
understanding of these kinds of reversible flowchart
languages~\cite{YokoyamaEtAl:2015,GlueckKaarsgaard:2018}.

Reversible functional programming languages are still at an early stage of
development, and today only a few proof-of-concept languages exist. This work
is founded on the initial work on
\rfun~\cite{YokoyamaAxelsenGlueck:2012:LNCS,ThomsenAxelsen:2016:IFL}, while
another notable example of a reversible functional language is
Theseus~\cite{JamesSabry:2014:RC}, which has recently been further developed
towards a language for quantum computations~\cite{SabryEtal:2018}.

%% Linear types

The type system formulated here is based on relevance logic (originally
introduced in \cite{Anderson:1975}, see also \cite{Dunn:2002}), a substructural
logic similar to linear logic~\cite{Girard:1987,Wadler:1990} which (unlike
linear logic) permits the duplication of data. In reversible functional
programming, linear type systems (see e.g.~\cite{JamesSabry:2014:RC}) have
played an important role in ensuring reversibility, but they also appear in
modern languages like the Rust programming language~\cite{Matsakis:2014}. To
support ancillary variables at the type level, we adapt a type system inspired
by Polakow's combined reasoning system of ordered, linear, and unrestricted
intuitionistic logic~\cite{Polakow:2001}.

The rest of this thesis is organised in the following way: In
Sect.~\ref{sec:formal} we will first introduce \rfunc followed by the type
system and operational semantics. We also discuss type polymorphism and show
that the language is indeed reversible. In Sect.~\ref{sec:staticFMP} we will
show how the type system in some cases can be used to statically verify the
first match policy. In Sect.~\ref{sec:running_backwards} we discuss program
inversion and present inverse operational semantics. Sect.~\ref{sec:prog} we
show how syntactic sugar can be used to design a more modern style functional
language from \rfunc. In Sect.~\ref{sec:implementation} we briefly introduce a
reference implementation. In Sect.~\ref{sec:discussion} we discuss language
design and future work. Finally in Sect.~\ref{sec:conclusion} we conclude.

%-------------------------------------------------------------------------------
% LANGUAGE
%-------------------------------------------------------------------------------

\input{language}

%-------------------------------------------------------------------------------
% FIRST MATCH GUARANTEE
%-------------------------------------------------------------------------------

\input{firstmatch}

%-------------------------------------------------------------------------------
% INVERSE
%-------------------------------------------------------------------------------

\input{inverse}

%-------------------------------------------------------------------------------
% PROGRAMMING
%-------------------------------------------------------------------------------

\input{programming}

%-------------------------------------------------------------------------------
% IMPLEMENTATION
%-------------------------------------------------------------------------------

\input{implementation}

%-------------------------------------------------------------------------------
% DISCUSSION
%-------------------------------------------------------------------------------

\input{discussion}

%-------------------------------------------------------------------------------
% CONCLUSION
%-------------------------------------------------------------------------------

\chapter{Conclusion}\label{sec:conclusion}

Although \rfunc is a continuation of the work that was started with \rfun, its
abstract syntax and evaluation semantics are quite different and include more
explicit primitive language constructs. However, we have also shown that
\rfunc can be made lighter via syntactic sugar to mimic other functional
languages.

We have presented a formal type system for \rfunc, including support for
recursive types through a fix point operator and polymorphic types via
parametric polymorphism. The type system is built on relevance typing, which is
sufficient for reversibility if we accept that functions may be partial.

Evaluation has been presented through a big step semantics. Most evaluation
rules were straightforward, but it was necessary to define a notion of leaves
and a relation for ``unification'' as machinery to describe the side condition
necessary to capture the first match policy.

An advantage offered by the type system is the ability to check the first match
policy statically. A static guarantee that the first match policy should hold
for a function will eliminate the run time overhead of case-expressions, often
leading to more efficient evaluation. By the program syntax, we can check for
orthogonality of inputs and the possible values of leaf expressions. By the
type system, we can argue for the finiteness of a function's domain and exhaust
the possible computational paths for a case-expression. Further, as was shown
in future work, we can apply an induction principle for recursive types.
However, it is difficult to detect exactly when this will yield a first match
policy guarantee.

We noted that, in contrast to many reversible programming languages, the syntax
of \rfunc does not easily support generation of inverse programs. This is not
problematic as the relational semantics do make it possible to inverse
interpret a program. We presented an inference system for inverse evaluation
and showed how inverse application of functions is achieved.

Finally, we have argued that it is possible to enhance the syntax of \rfunc
with high level constructs, which in turn have simple translation schemes back
to the core language. We have presented numerous examples: variants, type
classes (which, as an example, can be used to replace the duplication/equality
operator in the original \rfun language), top-level cases, records, arbitrarily
sized products and abbreviated let-expressions.

%-------------------------------------------------------------------------------
% BIBLIOGRAPHY
%-------------------------------------------------------------------------------

\bibliographystyle{splncs03}
\bibliography{references}

\end{document}
